% One-page project status handout for presentations (Health-JEPA).
% Build: pdflatex COHORT_COMPASS_STATUS_ONE_PAGE.tex
\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{amsmath,amssymb}
\usepackage{url}
\geometry{margin=0.6in}
\setlength{\parindent}{0pt}
% Extra space between paragraphs
\setlength{\parskip}{0.65em}
% Comfortable line leading within paragraphs
\onehalfspacing
\setlist[itemize]{nosep,leftmargin=1.2em,itemsep=0.4em,topsep=0.35em}
\newcommand{\sectiontitle}[1]{%
  \vspace{0.55em}\noindent\textbf{#1}\par\vspace{0.35em}}
\begin{document}
\thispagestyle{empty}
{\centering\large\textbf{Health-JEPA --- Project status (concise)}\\[0.35em]
\small Action-conditional JEPA for clinical time series + platform roadmap\par}
\vspace{0.55em}
\rule{\linewidth}{0.4pt}
\vspace{0.45em}

\sectiontitle{1.\ Overall summary}
\emph{Health-JEPA} is a \textbf{causal healthcare analytics platform}. The core is a
\textbf{Joint-Embedding Predictive Architecture (JEPA)} on irregular clinical time series:
it learns a latent representation of a patient's \emph{pre}-intervention trajectory and
predicts a \emph{post}-intervention representation under a discrete intervention $z$---%
\textbf{in embedding space} (MSE on latent vectors), with \textbf{no} reconstruction loss on raw labs.
Downstream, \textbf{vector search} (Qdrant ``twins''), an LLM/RAG layer, a \textbf{Next.js} UI,
and a \textbf{FastAPI + Celery} pipeline deliver narratives and a neighborhood visualization
for end-to-end analysis.

\sectiontitle{2.\ All of Us (AoU) data collection (work completed)}
A Researcher Workbench pipeline was implemented and run on the AoU OMOP CDR. The
\textbf{index event} is the \textbf{first drug exposure} (OMOP \texttt{drug\_exposure},
rolled up to ingredient level via \texttt{concept\_ancestor}). A fixed cohort cap per arm
was used to keep jobs tractable. \textbf{Pre/post windows} of $\pm 365$ days were extracted;
signals were \textbf{binned weekly} to reduce size and memory.

\textbf{EHR:} a curated set of \textbf{labs and vitals} (LOINC-aligned measurement concepts)---%
cardiometabolic and renal focus (e.g.\ HbA1c, lipids, creatinine, eGFR, blood pressure, BMI, SpO$_2$).
\textbf{Wearables:} daily Fitbit tables (\texttt{activity\_summary}, \texttt{heart\_rate\_summary},
and related summaries) were aggregated to \textbf{weekly} means aligned to the same windows.

\textbf{Training shard produced} (exemplar run): $K{=}3$ interventions
(metformin, atorvastatin, lisinopril), $F{=}25$ features (EHR $+$ wearable channels),
$N\approx 4.3$k valid patient--intervention examples with \textbf{roughly balanced} arm counts;
exports: \texttt{patient\_tensors.npz}, \texttt{feature\_vocab.json},
\texttt{intervention\_map.json}, and \texttt{manifest.json} for reproducibility.
\textit{Note:} observational EHR and optional wearables imply \textbf{missingness} handled via masks;%
claims are \textbf{associational} without an external causal ID strategy.

\sectiontitle{3.\ JEPA model (three sub-networks) and next steps}
The implementation (\texttt{CausalJEPA} / \texttt{ClinicalJEPA} alias) is an \textbf{I-JEPA-style}
\textit{three-module} system:

\begin{itemize}
  \item \textbf{Context encoder:} time-series encoder (value $+$ mask, continuous time) $\to$ pooled
        embedding $s_x$. \textbf{Trained} with backprop.
  \item \textbf{Target encoder:} same architecture; produces $s_y$ from the post window.
        \textbf{No gradients} from the loss; weights follow the context encoder by \textbf{EMA} only
        (momentum update each step).
  \item \textbf{Predictor + action embedding:} $\mathrm{embed}(z) \in \mathbb{R}^{z_{\mathrm{dim}}}$ and
        an MLP on $[s_x \,\Vert\, \mathrm{embed}(z)]$ outputs $\hat s_y$; training minimizes
        $\lVert \hat s_y - s_y \rVert^2$ with $s_y$ stopped.
\end{itemize}

\vspace{0.25em}
\noindent\textbf{Next steps:} (i)~wire a full \textbf{training loop} on the exported AoU shards with
$F,K$ and \texttt{max\_seq\_len} set from the manifest; (ii) save a \textbf{checkpoint} and align
inference \texttt{num\_interventions} with the \texttt{intervention\_map}; (iii) integrate
\textbf{twin verification} and product APIs for deployment tests.

\end{document}
